---
title: "01-cleaning"
format: html
---

## Goals of this notebook

The steps we'll take to prepare our data:

- Download the all data
- Import it into our notebook
- Build a vector for PEN data
- Remove unneeded columns
- Clean dates and column names
- Export the all clean data for next notebook

## Setup

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
```

## Import Data

Importing all files separately to view the data individually. Want to make sure all rows and columns are the same. 

PEN 2021-2022

```{r}
PEN_21 <- read_csv("data-original/laura-PEN/2021-2022_PEN_Index.csv")
```
PEN 2022-2023

```{r}
PEN_22 <- read_csv("data-original/laura-PEN/2022-2023_PEN_Index.csv")
```

PEN 2023-2024

```{r}
PEN_23 <- read_csv("data-original/laura-PEN/2023-2024_PEN_Index.csv")

PEN_23 |> glimpse()
```

PEN 2024-2025

```{r}
PEN_24 <- read_csv("data-original/laura-PEN/2024-2025_PEN_Index.csv")
```

District

```{r}
district_raw <- read_csv("data-original/laura-PEN/2024-2025_Districts.csv")
```

Statewide

```{r}
state_raw <- read_csv("data-original/laura-PEN/2024_2025_Statewide.csv")
```

## PEN data
Build a vector of files to import

```{r}
PEN_files_raw <- list.files(
  "data-original/laura-PEN",
  pattern = "PEN",
  full.names = TRUE
)

PEN_files_raw 
```

## Map, import, bind

```{r}
PEN_data <- PEN_files_raw |> 
  set_names(basename) |> 
    map(read_csv) |> 
  list_rbind(names_to = "school_year") |> 
  clean_names()

PEN_data |> nrow()
PEN_data |> head()
PEN_data |> tail()
```
```{r}
PEN_data |> 
  filter(school_year == "2023-2024_PEN_Index.csv")
```

## Clean school years, remove string column

```{r}
PEN_columns <- PEN_data |> 
  mutate(year = str_extract(school_year, "\\d{4}-\\d{4}")) |> 
  select(!school_year)

PEN_columns |> 
  arrange(desc(year)) 
```

## Join Ban Status and Ban Type column; Series and Series Name

```{r}
PEN_ban_type <- PEN_columns |> 
  mutate(ban_standing = if_else(is.na(type_of_ban), ban_status, type_of_ban)) |> 
  select(!c(ban_status, type_of_ban)) 
```

```{r}
PEN_files_clean <- PEN_ban_type |> 
  mutate(book_series = if_else(is.na(series), series_name, series)) |> 
  select(!c(series, series_name))
```

## Export Data

```{r}
PEN_files_clean |> 
  write_rds("data-processed/PEN-files.rds")
```

## State Data

```{r}
state_clean <- state_raw |> 
  clean_names()

state_clean
```

## Clean Dates

```{r}
state_dates <- state_clean |> 
  mutate(
    date_removed_challenged = ymd(date_of_challenge_removal)
  ) 

state_dates
```

## Remove Columns

```{r}
clean_states <- state_dates |> 
  select(!c(secondary_author_s,illustrator_s,translator_s,series,date_of_challenge_removal))
```

## Export Clean Data

```{r}
clean_states |> 
  write_rds("data-processed/statewide-ban.rds")
```

## District Data

```{r}
district_clean <- district_raw |> 
  clean_names()

district_clean
```

## Add a date column

```{r}
district_dates <- district_clean |> 
  mutate(
    date_removed = my(date_of_challenge_removal)
  )

district_dates 
```

## Export clean data

```{r}
district_dates |> 
  write_rds("data-processed/districts-bans.rds")
```

## Census data
Import census data
```{r}
census_raw <- read_csv("data-original/census-data.csv") 

census_raw
```

## Clean names, remove unnecessary columns

```{r}
census_clean <- census_raw |> 
  clean_names() |> 
  select(!c(b01003001_error,geoid))

census_clean
```

## Change column names

```{r}
avg_populations <- census_clean |> 
  mutate(state = name,
         population = b01003001) |> 
  select(!c(b01003001, name))

avg_populations
```

## Export data
```{r}
avg_populations |> 
  write_rds("data-processed/avg_populations.rds")
```
