---
title: "02-analysis"
format: html
---

## Set Up
```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
```


## Goals of this analysis, find the answers to the following questions:

* Which author is banned the most across all districts?
  + During each year?
  + In each state?
* Which state has the most banned titles overall?
* Which book most the most commonly banned across all districts?
  + During each school year?
+ Which districts in Texas banned the most books each year? The least?
+ Which districts in Texas have banned the most books overall? The least?
* How do the amount of bans compare to average population sizes?

  
## Import data
```{r}
PEN_files <- read_rds("data-processed/PEN-files.rds") 

PEN_files |> glimpse()
```

##  Which author has the most banned books across all districts? 

```{r}
PEN_files |> 
  group_by(title)
```

Data Takeaway:

Ellen Hopkins, Sarah J. Mass and Stephen King have been the top three most commonly banned authors between the 2021-22 and 2024-25 school years. Ellen Hopkins has the most banned titles, with 435 of her books banned across school districts nationwide. Sarah J. Mass has 338 titles banned, and Stephen King has 219 titles banned across school districts. 

#### Per year

```{r}
PEN_files |>
  group_by(author, year) |> 
  summarise(author_count = n()) |> 
  mutate(author_count = author_count |> as.numeric()) |> 
  filter(author_count >= 25) |> 
  arrange(year |> desc(), author_count |>  desc()) 
```

Data Takeaway:

Elle Hopkins had the most books banned in all school years, except for 2024-2024. That year, the most banned books where Stephen King titles, reaching a total of 206. During the 2023-2024 school year, 523 Elle Hopkins titles were banned, which is the most titles banned from a single author in a school year.

#### Per state 

```{r}

PEN_files |> 
  group_by(state) |> 
  count(author) |> 
  arrange(n |> desc())
```

Data Takeaway:

The most commonly banned author overall, Elle Hopkins, is also the most banned author in Iowa, where she has the most bans of anyone in a particular state. Iowa, Florida and Texas have a high proportion of bans compared to other states. In Texas, the most banned author is Elle Hopkins. Her books have been banned 117 times in the state since 2021. 

## Which state had the most banned titles during 2024-2025 school year

```{r}
PEN_files |> 
  filter(year=="2024-2025") |> 
  distinct(state, title, author) |> 
  group_by(state) |> 
  summarise(total_titles = n()) |> 
  arrange(desc(total_titles)) 
```

Data Takeaway: 

Overall, Florida has banned 4232 titles, which is greater than any other state. Texas ranks in second with 2987 banned titles.

## Which book was the most commonly banned across all districts? What about during a particular school year?

```{r}
PEN_files |> 
  group_by(author,title,year) |> 
  count(title, sort = T) 
```

```{r}
PEN_files |> 
  group_by(author,title) |> 
  count(title, sort = T) |> 
  mutate(n = n |> as.numeric()) |> 
  mutate(summed_appearances = n) |> 
  select(!n) |> 
  arrange(summed_appearances |> desc())
```

Data Takeaway: 

During the 2023-24 school year, "Nineteen Minutes" by Jodi Picoult was the most banned book. The book received 98 bans, more than any single book received in any other school year. That year, "Looking for Alaska" by John Green was banned 97 times. Overall, the 2023-24 school year saw the highest proportion of book bans for individual titles, including classics such as "The Handmaid's Tale" and "The Kite Runner."

## How many books were banned each state during each school year?
  
```{r}
PEN_files |> 
  group_by(state, year) |> 
  summarise(title = n()) |> 
  arrange(title |> desc())
```

Data Takeaway:

Overall, the most states that most consistently banned the highest amount of books were Florida and Texas. Iowa had the second highest amount of bans during the 2023-2024 school year with 3671 books banned. This was also the second highest number of bans in a single school year overall. Texas had the more bans than any other state during a single school year with 4561 bans during the 2023-2024 school year.

## Which districts in Texas banned the most books each year? The least?

```{r}
Texas_districts <- PEN_files |> 
  filter(state == "Texas") |> 
  group_by(district, year) |> 
  summarise(title = n()) |> 
  arrange(title |> desc()) 

Texas_districts
```

Data Takeaway:

North East Independent School District had more bans than any other district during the 2024-2025 school year with 752 titles banned. Many districts had years with only one book banned including Jacksonville Independent School District, Lake Travis Independent School District, Waller Independent School District and Highland Park Independent School District. Highland Park Independent School District was the only district in Texas to ban one book during the 2024-2025 school year.


## How do the amount of bans compare to average population sizes

Import data

```{r}
avg_populations <- read_rds("data-processed/avg_populations.rds")

avg_populations |> glimpse()
```

Filter for year
```{r}
ban_per_year <- PEN_files |> 
  filter(year == "2022-2023") |> 
  group_by(state) |> 
  summarise(total_titles = n()) |> 
  arrange(total_titles |> desc())

ban_per_year
```

Join columns
```{r}
titles_per_pop <- right_join(avg_populations, ban_per_year, by = "state") |> 
  arrange(total_titles |> desc())

titles_per_pop
```

Data Takeaway:
During the 2022-2023 school year, Florida banned more books than any other state with 1406 bans. This is more than double the amount books Texas banned during that year. Districts across Texas banned 625 books that year, despite the state having a population with nearly 8 million more people according to the 2019-2023 census data.

```{r}
titles_per_pop |> 
  group_by(state) |> 
    summarise(bans_per_pop = mean(total_titles / population * 100000, na.rm = TRUE)
  ) |> 
  arrange(bans_per_pop |> desc())

options(scipen = 999)

titles_per_pop |> 
  mutate(total_titles / (population * 100000))
```

There around around 8 books banned for ever 100,000 people in Utah. Whereas in Florida, the state with the most overall banned titles, there are 6 bans per 100,000 people. This ratio drastically decreases for states such including New York, New Jersey and Georgia, where there around .03 books banned per 100,000 people. This equates to 3 books being banned for every 1,000 people. In California, there are as few as .002 books banned per 100,000 people. This is equates to .2 books being banned for every 1,000 people. In Texas and Florida, the states with the two highest total books banned, the ratios per 1,000 are .021 book per 1,000 people and .064 books banned per 1,000 people, respectively. 

When considering population size states such as California, Massachusetts and Minnesota had the highest proportion of books removed from school districts. 

## Thought: what does population size really matter in this, it should more be comparing school districts per state to bring context to the overall number

install packages
```{r}
library(readxl)
```

import data

```{r}
read_excel("data-original/school_districts.xlsx")
```
pivot data???
and clean names


OR

```{r}
district_populations_raw <-
  read_csv("data-original/district_demographics.csv")

district_populations_raw
```

## Chart which school district had the most banned titles during the recent school year

```{r}
banned_titles <- PEN_files |> 
  filter(year=="2024-2025") |> 
  distinct(state, title, author) |> 
  group_by(state) |> 
  summarise(total_titles = n()) |> 
  filter(state!= "Nation") |> 
  arrange(desc(total_titles)) |> 
  head(10) 

banned_titles
```

Mutate total titles to be a numerical value
```{r}
banned_titles_chart <- banned_titles |> 
  mutate(
    total_titles = as.numeric(total_titles)
  )
```


```{r}
books_chart <- 
  banned_titles_chart |> 
  ggplot(
    aes(x = total_titles, y = reorder(state, total_titles))
  ) +
  geom_col() +
  geom_text(
    aes(
      label = total_titles,
      hjust = ifelse(total_titles > 500, 1.1, -0.1)
## According to ChatGPT, I can use "ifelse" to select which rows I want to use hjust for. 1.1 places the label slightly left of the bar end, while -0.1 places it slightly right of the bar end
    ),
    size = 3
  ) +
  labs(
    title = "The total of banned books is highest in Southern states",
    subtitle = str_wrap("Texas, Tennesee and Florida banned the most books during the 2024-2025 school year. These states banned over 1,000 books across their school districts, while other states banned less than 100."),
    x = "Banned Books", y = "State",
    caption = "By Laura Baker"
  ) +
  theme_classic() 

ggsave("figures/books_chart.png")
```


## Chart the top banned authors during 2024-2025 school year
```{r}
banned_authors_chart <- PEN_files |>
    filter(year == "2024-2025") |> 
  group_by(author, year) |> 
  summarise(author_count = n()) |> 
  mutate(author_count = author_count |> as.numeric()) |> 
  arrange(year |> desc(), author_count |>  desc()) |> 
    head(10) 

banned_authors_chart
```

```{r}
author_chart <- 
banned_authors_chart |> 
 ggplot(aes(x = author_count, y = reorder(author, author_count))) +
  geom_col() +
  labs(
    title = "The authors most commonly banned in school districts",
    subtitle = str_wrap("Fiction writers have been the most frequently banned in school districts. Stephen King, Ellen Hopkins and Sarah J. Maas are banned more than any others. These writers have had titles banned at least 100 times since 2022. This chart includes the ten authors with the most banned books."),
    x = "Books Banned", y = "Author",
    caption = "By Laura Baker"
  ) +
  
  theme_light()

ggsave("figures/author_chart_clean.png")
```

